**The status of the microbial census: an update**

### 2015-11-10 Schloss_Census2_mBio_2015.Rmd
Fill in Introduction

@@

## Introduction

The effort to quantify the number of different organisms in a system remains fundamental to understanding ecology. At the scale of microorganisms, small physical sizes, morphological ambiguity, and highly variable population sizes complicate this process immensely. Furthermore, creating standards for delimiting what makes one microbe "different” from another has proved contentious at best. In spite of these challenges, we continue to peel back the curtain on the microbial world with the aid of more and more informative, if still limited, technologies like cultivation, 16S rRNA gene surveys, and metagenomics.

In 2004, Schloss and Handelsman analyzed 56,215 partial rRNA gene sequences available in the Ribosomal Database Project and concluded that "either current sampling methods are not adequate to identify 10<sup>7</sup> to 10<sup>9</sup> different species or these estimates are high (Schloss, 2004)." Since that time, the explosion of short-read high throughput sequencing technology has caused a dramatic increase in the number of sequences deposited in public databases. Currently there are over XXXXXX 16S rRNA gene sequences in SRA. Recent estimates based on this short read data place the total number of microbial species at around XXXXX.

However, the increase in sequencing volume has come at the cost of sequence length. The commonly used MiSeq-based sequencing technology amplifies the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are shorter than 500 bp. The decrease in length means a decrease in informative sites, making direct comparisons between short read sequencing and mostly or completely full-length sequences difficult. Youssef et al. (2009) demonstrated high variability between the number of OTUs that different hypervariable sites estimated compared to full-length 16S rRNA gene sequences, and also that sample location influenced this behavior. Thus, it remains unclear to what degree richness estimates from short read technology over or underestimate the numbers from full-length sequences.

Here we update the status of the microbial census with nearly or completely full-length 16S rRNA gene sequences. In the 12 years since the collection of data for Schloss and Handelsman’s analysis, the number of available sequences has risen over twenty-fold, despite the overwhelming contemporary focus by most researchers on short-read technologies. This update to the census allows us to evaluate the relative sampling thoroughness for different environments and clades, and make an argument for the continued need to collection full-length sequence data from many systems that have a long history of study.

* Lack of references

@@

***Metadata curation.***
The environmental origins of the 16S rRNA gene sequences were manually classified using seven broad "coarse” categories, and further refined to facilitate additional analyses with twenty-six more specific "fine” categories (Table 1). These were assigned based on manual curation of the `isolation_source` category within the ARB database associated with each of the sequences. There were 20,084 bacterial and 1,613 archaeal isolation_source terms that were manually assigned to the coarse and fine categories in Table 1. For source definitions that were identifiable by online searches, educated guesses were made or they were placed into the coarse "Other” category. There were 150,378 bacterial and 2,604 archaeal sequences where an `isolation_source` term was not collected. Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our categorization are available at [@figshare???].

***Data analysis.***
Our analysis made use of ARB [OS X v.6.0, @Pruesse2007], mothur mothur [v.1.37.0; @Schloss2009], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015]. Within R we utilized knitr [v.`r packageDescription("knitr")$Version`; @xie2013dynamic] and openxlsx [v. `r packageDescription("openxlsx")$Version`; @Walker2015] packages. A reproducible version of this manuscript including data extraction and processing is available at https://www.github.com/SchlossLab/Schloss_Census2_XXXX_2015.

@@

>>>>

Importance goes here.

@@

## Introduction
In 1983, the full-length 16S rRNA gene sequence of *Escherichia coli* (accession J01695) was deposited into NCBI's GenBank making it the first of more than 10 million 16S rRNA gene sequences to be deposited into the database [@Brosius1978]. GenBank accessions represent nearly one-third of all sequences deposited in the database making it the best-represented gene. As Sanger sequencing has given way to to-called "next generation sequencing" technologies, hundreds of millions of 16S rRNA gene sequences have been deposited into the NCBI's Sequence Read Archive. The expansion in sequencing throughput and increased access to sequencing technology has allowed for more environments to be sequenced at a deeper coverage resulting in the identification of novel taxa. The ability to obtain sequence data from microorganisms without cultivation has radically altered our perspective of their role in nearly every environment from deep ocean sediment cores (e.g. accession AY436526) to the International Space Station (e.g. accession DQ497748). The effort to quantify the number of different organisms in a system remains fundamental to understanding ecology [@reference]. At the scale of microorganisms, small physical sizes, morphological ambiguity, and highly variable population sizes complicate this process. Furthermore, creating standards for delimiting what makes one microbe "different” from another has been contentious [@reference]. In spite of these challenges, we continue to peel back the curtain on the microbial world with the aid of more and more informative, if still limited, technologies like cultivation, 16S rRNA gene surveys, and metagenomics.

The increase in sequencing volume has come at the cost of sequence length. The commonly used MiSeq-based sequencing platform from Illumina is widely used to sequence the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are generally shorter than 500 bp. Perhaps most disconcerting about this development is the sense that the increased read depth is being gained using short read platforms rather than the full-length sequences. Because these short reads are used for classification to existing taxa, we are missing the opportunity to propose novel candidate taxa and vastly underappreciating the biodiversity of microbial life. We likely lack the references necessary to adequately classify the novel biodiversity we are sampling when we generate 100-times the sequence data from a community than we did using full-length sequencing.

Previously, Schloss and Handelsman [-@Schloss2004] assigned the then available 56,215 partial rRNA gene sequences to operational taxonomic units (OTUs) that were available in the Ribosomal Database Project and concluded that the sampling methods of the time were insufficient to identify the previously estimated 10^7^ to 10^9^ different species [@Dykhuizen1998, @Curtis2002]. That census called for a broader and deeper characterization of all environments. Refreshingly, this challenge was largely met. There have been major investments in studying the Earth's microbiome using 16S rRNA gene sequencing through initiatives such as the Human Microbiome Project [@reference], the Earth Microbiome Project [@reference], and the International Census of Marine Microorganisms. But most important, the original census was performed on the cusp of radical developments in sequencing technologies. That advancement has largely moved the bulk of sequencing throughput from large sequencing centers to individual investigators and leveraged their diversity to expand the representation of organisms and environments represented in public databases.

Here we update the status of the microbial census with nearly or completely full-length 16S rRNA gene sequences. In the 13 years since the collection of data for Schloss and Handelsman’s analysis, the number of full-length sequences has grown exponentially, despite the overwhelming contemporary focus by most researchers on short-read technologies. This update to the census allows us to evaluate the relative sampling thoroughness for different environments and clades, and make an argument for the continued need to collection full-length sequence data from many systems that have a long history of study. Although there has been a robust growth in the number of full-length sequences deposited to GenBank since its creation in 1983, the rate of growth has stalled over the past 5 years and the deposits have been dominated by a handful of research groups studying a limited number of environments. As researchers consider coalescing into a Unified Microbiome Initiative [@Alivisatos2015], it will be important to balance the need for mechanism-based studies with the need to generate full-length reference sequences from a diversity of environments.

@@

***Metadata curation.***
The environmental origins of the 16S rRNA gene sequences were manually classified using seven broad "coarse” categories, and further refined to facilitate additional analyses with twenty-six more specific "fine” categories (Table 1). These were assigned based on manual curation of the `isolation_source` category within the ARB database associated with each of the sequences. There were 20,084 bacterial and 1,613 archaeal isolation_source terms that were manually assigned to the coarse and fine categories in Table 1. For source definitions that were identifiable by online searches, educated guesses were made or they were placed into the coarse "Other” category. There were 150,378 bacterial and 2,604 archaeal sequences where an `isolation_source` term was not collected. Complete tables containing the ARB-provided metadata, taxonomic information, OTU assignment, and our categorization are available at [@figshare].

***Data analysis.***
Our analysis made use of ARB [OS X v.6.0, @Pruesse2007], mothur mothur [v.1.37.0; @Schloss2009], and R [v.`r paste(version$major, version$minor, sep='.')`; @language2015]. Within R we utilized knitr [v.`r packageDescription("knitr")$Version`; @xie2013dynamic] and openxlsx [v. `r packageDescription("openxlsx")$Version`; @Walker2015] packages. A reproducible version of this manuscript including data extraction and processing is available at https://www.github.com/SchlossLab/Schloss_Census2_mBio_2015.

@@

The first 16S sequence was published in 1978, not deposited until 1983. A bit of an allegory for our time.
